Pillar 1 - Sequential Pattern Mining

### 2019-01-29 Pillar 1 - Sequential Pattern Mining.Rmd

# Student level
d_transcript %>%
  group_by(`Student ID`) %>%
  provide_statistics()

# Course level
d_transcript %>%
  inner_join(d_course, by = c("Course ID")) %>%
  group_by(`Course ID`) %>%
  provide_statistics()

# Cluster level
d_transcript %>%
  inner_join(d_course, by = "Course ID") %>%
  filter(!is.na(Cluster)) %>%
  group_by(Cluster) %>%
  provide_statistics()

# Concentration evel
d_transcript %>%
  inner_join(d_course, by = "Course ID") %>%
  gather(X, Concentration, Concentration, `Concentration (additional)`, na.rm = TRUE) %>%
  group_by(Concentration) %>%
  provide_statistics()

# Year level
d_transcript %>%
  group_by(Year_numerical) %>%
  provide_statistics()

# Level level
d_transcript %>%
  inner_join(d_course, by = c("Course ID")) %>%
  filter(!is.na(Level)) %>%
  group_by(Level) %>%
  provide_statistics()
```

@@

## Creating Transactions
First we transform our data into transaction data. For this, we first create a vector of mandatory courses that we exclude from transcripts.

@@

#
# Threshold: low vs high grade
low_grade <- 6.5

@@

  # Identifying item
  mutate(
    PF = case_when( Grade <  5.5 ~ "fail",
                    Grade >= 5.5 ~ "pass"),
    HL = case_when( Grade <  low_grade ~ "low",
                    Grade >= low_grade ~ "high"),

@@

### Transactions for Apriori
```{r AR data prep 2}
transactions <- d_transactions %>%
  group_by(
    sequenceID
    ) %>%
  summarize(
    list_item = list(item)
    ) %>%
  ungroup

transactions <- as(transactions$list_item, "transactions")
```

```{r AR data prep 3 (PF)}
transactions_PF <- d_transactions %>%
  group_by(
    sequenceID
    ) %>%
  summarize(
    list_item = list(item_PF)
    ) %>%
  ungroup

transactions_PF <- as(transactions_PF$list_item, "transactions")
```

```{r AR data prep 4 (HL)}
transactions_HL <- d_transactions %>%
  group_by(
    sequenceID
    ) %>%
  summarize(
    list_item = list(item_HL)
    ) %>%
  ungroup

transactions_HL <- as(transactions_HL$list_item, "transactions")
```

### Sequences for CSPADE
```{r}
sequences_temp <- d_transactions %>%
  group_by(
    sequenceID,
    eventID
    ) %>%
  summarize(
    list_item = list(item)
    ) %>%
  arrange(
    sequenceID,
    eventID
    )

sequences             <- as(sequences_temp$list_item, "transactions")
sequences@itemsetInfo <- as.data.frame(select(sequences_temp, sequenceID, eventID))

#checking results
#inspect(head(sequences, 10))
```

```{r}
sequences_temp <- d_transactions %>%
  group_by(
    sequenceID,
    eventID
    ) %>%
  summarize(
    list_item = list(item_PF)
    ) %>%
  arrange(
    sequenceID,
    eventID
    )

sequences_PF             <- as(sequences_temp$list_item, "transactions")
sequences_PF@itemsetInfo <- as.data.frame(select(sequences_temp, sequenceID, eventID))

#checking results
#inspect(head(sequences_PF, 10))
```

```{r}
sequences_temp <- d_transactions %>%
  group_by(
    sequenceID,
    eventID
    ) %>%
  summarize(
    list_item = list(item_HL)
    ) %>%
  arrange(
    sequenceID,
    eventID
    )

sequences_HL             <- as(sequences_temp$list_item, "transactions")
sequences_HL@itemsetInfo <- as.data.frame(select(sequences_temp, sequenceID, eventID))

#checking results
#inspect(head(sequences_HL, 10))
```

## Apriori Algorithm
```{r}
clean_AR <- function(AR){

  AR %>%
    as("data.frame") %>%
    filter(
      count >= 1
    ) %>%
    mutate(
      rules = str_remove(rules, pattern = "_fail"),
      lhs = substr(rules, 2, 8),
      rhs = substr(rules, 15, 21),
      `rhs.support` = confidence / lift
      ) %>%
    mutate_at(
      c("support", "lhs.support", "confidence", "rhs.support", "lift"),
      funs(round(., 5))
      ) %>%
    select(
      lhs, rhs,
      support, count, lhs.support,
      confidence, rhs.support,
      lift
      ) %>%
    arrange(
      desc(count)
    )

@@

We apply Apriori algorithm:
```{r AR apriori}
AR <- apriori(
  data = transactions, 
  parameter = list(
    supp = 0,    # min support
    smax = 1,    # max support
    conf = 0,    # min confidence
    minlen = 2,  # min length of rule
    maxlen = 2,  # max length of rule
    ext = TRUE,
    arem = "diff",
    aval = TRUE,
    minval = 0
    ),
  appearance = NULL,
  control = list(
    verbose = FALSE
    )
  ) %>%
  clean_AR

@@

```{r apriori PF}
# creating vector of fail course
course_id_fail <- d_transactions_PF %>%
  filter(PF == "fail") %>%
  distinct(`Course ID`)

# Apply apriori
AR_PF <- apriori(
  data = transactions_PF, 
  parameter = list(
    supp = 0, # min support
    smax = 1,    # max support
    conf = 0,    # min confidence
    minlen = 2,  # min length of rule
    maxlen = 2,  # max length of rule
    ext = TRUE,
    arem = "diff",
    aval = TRUE,
    minval = 0
    ),
  appearance = list(
    both = course_id_fail$`Course ID` # only include rules with fail on LHS and RHS
    ),
  control = list(
    verbose = FALSE
    )
  ) %>%
  clean_AR
```

```{r AR exploration}

data_rules <- AR
data_rules <- AR_PF

# Best support
data_rules %>%
  arrange(desc(support))

# Best confidence
data_rules %>%
  filter(count >= 5) %>%
  arrange(desc(confidence))

# Best lift
data_rules %>%
  filter(count >= 5) %>%
  arrange(desc(lift))
```

@@

```{r}
subsequences <- sequences %>%
  cspade(
    parameter = list(
      support = 0, # min support of a sequence
      maxsize = 1, # max number of items of an element of a sequence
      maxlen  = 2, # max number of element of a sequence
      mingap  = 1, # min time difference between consecutive element of a sequence
      maxgap  = 1e4, # max time difference between consecutive element of a sequence
      maxwin  = 1e4  # max time difference between any two elements of a sequence
      ),
    control = list(
      verbose = FALSE
      )
    )

@@

#EXPERIMENTS:
##Condensing grades

@@

@@

distributed_transcript <- spread(tmp_distributed_transcript, key = "Round_grade", value = Values #, fill = F
               )%>%
  mutate(g1  = g0|g1,
         g2  = g1|g2,
         g3  = g2|g3,
         g4  = g3|g4,
         g5  = g4|g5,
         g6  = g5|g6,
         g7  = g6|g7,
         g8  = g7|g8,
         g9  = g8|g9,
         g10 = g9|g10,
         ) %>%
  gather(key="Gr_string", value="Gr_bool", g0, g1, g2, g3, g4,g5,g6,g7,g8,g9,g10, na.rm=T) %>%
  select(-Gr_bool )%>%
  separate(Gr_string, into=c("Trash", "Grade_new"), sep="g")%>%
  select(-Trash) %>%
  mutate(Grade= as.numeric(Grade_new))%>%
  select(-Grade_new)%>%
  filter(!str_detect(Period, " to "))%>% #Crutch FILTER
  mutate(Item=paste(`Course ID`,Grade, sep="_"))%>%
  mutate( Time = paste(Year_numerical, Period, sep = ""),
          Time = as.double(Time))%>%
  select(`Student ID`, Item, Time) %>%
  arrange(`Student ID`, Time) %>% #Note that integer identifiers must be positive and that transactions must be ordered by sequenceID and eventID.
  group_by(`Student ID`, Time) %>%
  summarize(list_course = list(Item)) %>%
  mutate(TransactionID = paste(`Student ID`, Time, sep = "_"))

@@

#checking results
as(transactions, "data.frame")
```
##Create filter condition.
```{r}
# string_filter <- case_when(quo_name(category)=="Outcome"        ~ "Fail",
#                            quo_name(category)=="Low_grade"      ~ "low score",
#                            quo_name(category)=="Discrete_grade" ~ "5",
#                            quo_name(category)=="None"           ~ "_" #useless filter
#                            )
```
##CSPADE Algorithm:

@@

```
here I'l playing around with the correct confidence:
#correcting lookup for this case:
First we find out raw proportions per course (how many were low grade/high grade, out of a particular course, and the percentage this represents from the total)
```{r}
lookup_table <- d_course %>%

@@

  # Compute statistics
  group_by(LHS,`RHS Course ID`) %>%
  arrange(LHS,`RHS Course ID`) %>%
  mutate(
    Prob_both = sum(support), 
    Conf_corr = support/Prob_both,
    n_rule    = support * n_students,
    n_both    = Prob_both * n_students,
    lift_diff = Conf_corr - Prop_low,
    lift_corr = Conf_corr / Prop_low
    ) %>%
  ungroup() %>%

@@

---------------------------------------------------

##Puting students into take and not take. 
```{r}
#This df has 731,676 rows:
test <- expand.grid(`Student ID`= unique(d_transcript$`Student ID`),`Course ID`= unique(d_course$`Course ID`)) %>%
  left_join(d_transcript, by=c("Course ID", "Student ID")) %>%
  select(`Course ID`, `Student ID`, Grade, Year_numerical, Period) %>%
  mutate(
    take_course = case_when(
      is.na(Grade)  ~ FALSE,
      T             ~ TRUE
      ),
    PFN = case_when(
      is.na(Grade)  ~ "not",
      Grade >= 5.5  ~ "pass",
      Grade < 5.5   ~ "fail"
      )
    )

@@

>>>>

# Student level
statistics_student <- d_transcript %>%
  group_by(`Student ID`) %>%
  provide_statistics()

# Course level
statistics_course <- d_transcript %>%
  inner_join(d_course, by = c("Course ID")) %>%
  group_by(`Course ID`) %>%
  provide_statistics()

# Cluster level
statistics_cluster <- d_transcript %>%
  inner_join(d_course, by = "Course ID") %>%
  filter(!is.na(Cluster)) %>%
  group_by(Cluster) %>%
  provide_statistics()

# Concentration evel
statistics_concentration <- d_transcript %>%
  inner_join(d_course, by = "Course ID") %>%
  gather(X, Concentration, Concentration, `Concentration (additional)`, na.rm = TRUE) %>%
  group_by(Concentration) %>%
  provide_statistics()

# Year level
statistics_year <- d_transcript %>%
  group_by(Year_numerical) %>%
  provide_statistics()

# Level level
statistics_level <- d_transcript %>%
  # TODO: filter for student who completed their studies
  inner_join(d_course, by = c("Course ID")) %>%
  filter(!is.na(Level)) %>%
  group_by(Level) %>%
  provide_statistics()

#
# output
save(statistics_student, statistics_course, statistics_cluster,
     statistics_concentration, statistics_year, statistics_level, file = "Output/Transcript Statistics.RDATA")
rm(provide_statistics, statistics_student, statistics_course, statistics_cluster,
     statistics_concentration, statistics_year, statistics_level)
```

@@

## Data Prep: Creating Transactions and Sequences
First we transform our data into transaction data. For this, we first create a vector of mandatory courses that we exclude from transcripts.

@@

#
# Threshold: pass grade, high grade
pass_grade <- 5.5
high_grade <- 6.5

@@

  # Identifying item
  mutate(
    PF = case_when( Grade <  pass_grade ~ "fail",
                    Grade >= pass_grade ~ "pass"),
    HL = case_when( Grade <  high_grade ~ "low",
                    Grade >= high_grade ~ "high"),

@@

### Transactions for Apriori
```{r making transactions}

#
# For convenience
make_transaction <- function(data = d_transactions, item = item){

  data <- data %>%
    group_by(
      sequenceID
    ) %>%
    summarise(
      list_item = list(!!enquo(item))
    ) %>%
    ungroup

  transactions <- as(
    data$list_item,
    "transactions"
    )

  return(transactions)

#
# Making transactions
transactions    <- make_transaction(item = item   )
transactions_PF <- make_transaction(item = item_PF)
transactions_HL <- make_transaction(item = item_HL)

#
# Checking transactions
inspect(head(transactions, 10))

rm(make_transaction)
```

### Sequences for CSPADE
```{r}

#
# For convenience
make_sequence <- function(data = d_transactions, item = item){

  data <- data %>%
    group_by(
      sequenceID,
      eventID
    ) %>%
    summarise(
      list_item = list(!!enquo(item))
    ) %>%
    arrange(
      sequenceID,
      eventID
    ) %>%
    ungroup

  sequences <- as(
    data$list_item,
    "transactions"
    )

  sequences@itemsetInfo <- select(
    data,
    sequenceID,
    eventID
    ) %>%
    as.data.frame()

  return(sequences)

#
# Making sequences
sequences    <- make_sequence(item = item   )
sequences_PF <- make_sequence(item = item_PF)
sequences_HL <- make_sequence(item = item_HL)

#
# Checking sequences
inspect(head(sequences, 10))

rm(make_transaction)
```

## Mining Rules

### Apriori Algorithm
```{r}

# Transform AR from apriori into a readable data frame
clean_AR <- function(AR){

  AR %>%
    as("data.frame") %>%
    filter(
      count >= 1
    ) %>%
    mutate(
      rules = str_remove(rules, pattern = "_fail"),
      rules = str_remove(rules, pattern = "_low"),
      lhs = substr(rules, 2, 8),
      rhs = substr(rules, 15, 21),
      `rhs.support` = confidence / lift
      ) %>%
    mutate_at(
      c("support", "lhs.support", "confidence", "rhs.support", "lift"),
      funs(round(., 5))
      ) %>%
    select(
      lhs, rhs,
      support, count, lhs.support,
      confidence, rhs.support,
      lift
      ) %>%
    arrange(
      desc(count)
    )

@@

We apply Apriori algorithm:
```{r AR apriori}

# Run the apriori algorithm with desired parameters
my_apriori <- function(data, appearance = NULL){

  data %>%
    apriori(
      parameter = list(
        supp = 0,    # min support
        smax = 1,    # max support
        conf = 0,    # min confidence
        minlen = 2,  # min length of 
        maxlen = 2,  # max length of rule
        ext = TRUE,
        arem = "diff",
        aval = TRUE,
        minval = 0
        ),
      appearance = appearance,
      control = list(
        verbose = FALSE
        )
      ) %>%
    clean_AR

#
# Generating asocciation rules
AR <- my_apriori(transactions)

# creating vector of fail course
course_id_fail <- d_transactions %>%
  filter(PF == "fail") %>%
  distinct(`item_PF`)

AR_PF <- my_apriori(transactions_PF,
                    appearance = list(both = course_id_fail$`item_PF`))

# creating vector of fail course
course_id_low <- d_transactions %>%
  filter(HL == "low") %>%
  distinct(`item_HL`)

AR_HL <- my_apriori(transactions_HL,
                    appearance = list(both = course_id_low$`item_HL`))

#
# Save association rules
save(AR, AR_PF, AR_HL,
     file = "Output/AS.RDATA")

rm(clean_AR, my_apriori,
   transactions, transactions_PF, transactions_HL,
   course_id_fail, course_id_low
   AR, AR_PF, AR_HL)

@@

@@

subsequences <- sequences %>%
  cspade(
    parameter = list(
      support = 0, # min support of a sequence
      maxsize = 1, # max number of items of an element of a sequence
      maxlen  = 2, # max number of element of a sequence
      mingap  = 1, # min time difference between consecutive element of a sequence
      maxgap  = 1e4, # max time difference between consecutive element of a sequence
      maxwin  = 1e4  # max time difference between any two elements of a sequence
      ),
    control = list(
      verbose = FALSE
      )
    )

@@

#EXPERIMENTS:

##TO MOVE UP (These are ready:)
Rounds grades and add new row from original grade up to 10.
NOTE: includes mandatory courses.
```{r}
distributed_transcript <- d_transcript %>%
  filter(!is.na(Grade), Grade>=0) %>%
  mutate(Round_grade = paste("g", round(Grade,0),sep=""),
         Values = T) %>%
  spread(key = "Round_grade", value = Values) %>%
  mutate(g1  = g0|g1,
         g2  = g1|g2,
         g3  = g2|g3,
         g4  = g3|g4,
         g5  = g4|g5,
         g6  = g5|g6,
         g7  = g6|g7,
         g8  = g7|g8,
         g9  = g8|g9,
         g10 = g9|g10,
         ) %>%
  gather(key="Gr_string", value="Gr_bool", g0, g1, g2, g3, g4,g5,g6,g7,g8,g9,g10, na.rm=T) %>%
  select(-Gr_bool )%>%
  separate(Gr_string, into=c("Trash", "Grade_new"), sep="g")%>%
  select(-Trash) %>%
  mutate(Grade= as.numeric(Grade_new))%>%
  select(-Grade_new) #%>%
  #to check that the grades appear from original grade to 10.
  #arrange(`Student ID`, `Course ID`) 
```

#TRASH:
##Condensing grades

@@

Excluding manatory courses.
```{r}

@@

distributed_transcript <- spread(tmp_distributed_transcript, key = "Round_grade", value = Values #, fill = F
               ) %>%
  mutate(g1  = g0|g1,
         g2  = g1|g2,
         g3  = g2|g3,
         g4  = g3|g4,
         g5  = g4|g5,
         g6  = g5|g6,
         g7  = g6|g7,
         g8  = g7|g8,
         g9  = g8|g9,
         g10 = g9|g10,
         ) %>%
  gather(key="Gr_string", value="Gr_bool", g0, g1, g2, g3, g4,g5,g6,g7,g8,g9,g10, na.rm=T) %>%
  select(-Gr_bool )%>%
  separate(Gr_string, into=c("Trash", "Grade_new"), sep="g")%>%
  select(-Trash) %>%
  mutate(Grade= as.numeric(Grade_new))%>%
  select(-Grade_new)%>%
  filter(!str_detect(Period, " to "))%>% #Crutch FILTER
  mutate(Item=paste(`Course ID`,Grade, sep="_"))%>%
  mutate( Time = paste(Year_numerical, Period, sep = ""),
          Time = as.double(Time))%>%
  select(`Student ID`, Item, Time) %>%
  arrange(`Student ID`, Time) %>% #Note that integer identifiers must be positive and that transactions must be ordered by sequenceID and eventID.
  group_by(`Student ID`, Time) %>%
  summarize(list_course = list(Item)) %>%
  mutate(TransactionID = paste(`Student ID`, Time, sep = "_"))

@@

#checking results
as(transactions, "data.frame")
```

##Create filter condition.
```{r}
# string_filter <- case_when(quo_name(category)=="Outcome"        ~ "Fail",
#                            quo_name(category)=="Low_grade"      ~ "low score",
#                            quo_name(category)=="Discrete_grade" ~ "5",
#                            quo_name(category)=="None"           ~ "_" #useless filter
#                            )
```
##CSPADE Algorithm:

@@

```
here I'l playing around with the correct confidence:

#correcting lookup for this case:
First we find out raw proportions per course (how many were low grade/high grade, out of a particular course, and the percentage this represents from the total)
```{r}
lookup_table <- d_course %>%

@@

  # Compute statistics
  group_by(LHS,`RHS Course ID`) %>%
  mutate(
    Prob_both = sum(support))%>%
  ungroup()%>%
  mutate( 
    Conf_corr = support/Prob_both,
    n_rule    = support * n_students,
    n_both    = Prob_both * n_students,
    lift_diff = Conf_corr - Prop_low,
    lift_corr = Conf_corr / Prop_low
    )

@@

##Puting students into take and not take. 
```{r}
#This df has 731,676 rows:
test <- expand.grid(`Student ID`= unique(d_transcript$`Student ID`),`Course ID`= unique(d_course$`Course ID`)) %>%
  left_join(d_transcript, by=c("Course ID", "Student ID")) %>%
  select(`Course ID`, `Student ID`, Grade, Year_numerical, Period) %>%
  mutate(
    take_course = case_when(
      is.na(Grade)  ~ FALSE,
      T             ~ TRUE
      ),
     PFN = case_when(
       is.na(Grade)  ~ "not",
       Grade >= 5.5  ~ "pass",
       Grade < 5.5   ~ "fail"
       ),
    Item = paste(`Course ID`, take_course, sep="_")
    )

@@

```{r}
test %>%
 mutate(
    Period  = substr(Period, 1, 1),
    eventID = as.numeric(paste(Year_numerical, Period, sep = "")),
    Item = paste(`Course ID`,Grade, sep="_")
    ) %>%
  select(`Student ID`, Item, eventID) %>%
  arrange(`Student ID`, eventID) %>% #Note that integer identifiers must be positive and that transactions must be ordered by sequenceID and eventID.
  group_by(`Student ID`, Time) %>%
  summarize(list_course = list(Item)) %>%
  mutate(TransactionID = paste(`Student ID`, Time, sep = "_"))
```
